---
title: "final_code"
author: "Sofia Perez"
date: "2026-03-16"
format: html
editor: visual
toc: true # use this to display a table of contents
---

github link:https://github.com/sofiaperez422/ENVS_193DS_winter-2026_final

## set up

```{r}
#| message: false
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(showtext) # to show custom text font on plots
library(MuMIn) # model selection

font_add_google(name = "Montserrat", # name of font on Google font website
              family = "mont") # name to call in font as
showtext_auto()

kelp <- read_csv(here("data",
                      "Annual_Kelp_All_Years_20250903.csv")) # save annual-kelp data as object  kelp
mopl <- read_csv(here("data",
                      "MOPL_nest-site_selection_Duchardt_et_al._2019.csv")) # save MOPL-nest data as object MOPL
personal_data <- read_csv(here("data", 
                      "Personal_Data20260315.csv")) # save personal data as object
```

# Problem 1.

## a.

In part 1, they used Linear Regression. In part 2, they used Kruskal-Wallis rank sum test.

## b.

To accompany part 2, they could make a box plot, where the x-axis would be water year classification and the y-axis would be the flooded wetland area (m^2^ ). The visual components would show the individual distributions of each water year classification with quartiles that reveal the medians, as well as outliers that may exist as single points.

## c.

Part 1 – The change in flooded wetland area is not predicted by precipitation (linear regression, F(degrees of freedom) = F statistic, R^2^ = coefficient of determination, p = 0.11, $\alpha$ = significance level).

Part 2 – We found no difference ($\eta^2$ = effect size) in median flooded wetland area across water year classification (wet, above normal, below normal, dry, critical drought) (Kruskal-Wallis test, $\chi^2$(degrees of freedom) = H statistic, p = 0.12, $\alpha$ = significance level).

## d.

An additional variable that could influence the flooded area of a wetland is the daily Delta outflow, specifically applied as a median daily, which would be a continuous variable. This might influence measured wetland flooding area because it directly involves the activity of water managers intentionally moving water on and off flooded wetland units for reasons like leaching salts from soils and providing wetland habitat for breeding waterfowl around that time of year, which has potential to influence the overall area.

# Problem 2.

## a.

```{r}
#| label: data cleaning kelp
#| warning: false
kelp_clean <- kelp |> # create new clean dataframe titled kelp_clean from kelp
  
# function 3
clean_names() |>  # clean column names to lowercase and underscores

  # function 1
filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) |>  # filter out observations from 2000-09-22 with transect 6 and quad 40

# function 6
filter(site %in% c("IVEE", "NAPL", "CARP")) |>  # filter to only include sites "IVEE" "NAPL" and "CARP"

# function 8
mutate(site_name = case_when( # rename sites to full site names
  site == "NAPL" ~ "Naples",
  site == "IVEE" ~ "Isla Vista",
  site == "CARP" ~ "Carpinteria"
)) |> 

# function 2
mutate(site_name = as_factor(site_name), # convert site name entries to factors
       site_name = fct_relevel(site_name, # reorder sites using factor properties
                               "Naples", "Isla Vista", "Carpinteria")) |> 

# function 5
group_by(site_name, year) |>  # group by site and year

# function 7
summarize(mean_fronds = mean(fronds)) |>  # calculate mean of fronds

# function 4
ungroup() # remove grouping

str(kelp_clean) # display the structure in a compact format
```

```{r}
slice_sample(kelp_clean, # display 5 random rows from the sample
             n = 5)
```

## b.

In the notes of the data set for that row it states "missing datasheet", and the number for the value of fronds in that "quad" in that "transect" is -99999. This value represents an "NA" value used when the data is either lost or not collected for a particular species (Methods and Protocols tab of the data page.)

## c.

```{r}
#| label: kelp visualization
ggplot(data = kelp_clean, # use kelp_clean data
       aes(x = year, # set x-axis
           y = mean_fronds, # set y-axis
           color = site_name)) + # group by site
  geom_line(linewidth = 0.8) + # plot lines and adjust thickness
  geom_point() + # plot points
  scale_color_manual(values = c("Isla Vista" = "bisque3", # change colors
                                "Naples" = "turquoise4", 
                                "Carpinteria" = "azure4")) + 
  theme_minimal(base_family = "mont") + # change theme and font
  labs(x = "Year", # label x-axis
       y = "Mean kelp fronds per year", # label y-axis
       title = "Sites vary in mean kelp fronds per year", # add plot title
       color = "Site") + # edit legend title
  theme(plot.title = element_text(hjust = 0.5)) # adjust title to center
```

## d.

The site with the most variable mean kelp fronds per year is Isla Vista, as it has the highest points compared to its lowest points seen with the longest lines. The site with the least variable mean kelp fronds per year is Carpinteria, as it has the least peaking relative to the rest of its data points with shorter lines and is also the most consistent in the y-axis region that the points stay in.

# Problem 3.

## a.

The 1s and 0s represent the Mountain Plover nest use, 1 meaning presence of nest use and 0 meaning no nest use.

## b.

Visual obstruction is a continuous variable that is metric incorporating both vegetation height and density, which are values that are measured with metric units. As visual obstruction increases the probability of Mountain Plover nest site use would decrease, and this is due to activity of black-tailed prairie dog colonies clipping vegetation to maintain visibility for predators, which makes for ideal nesting habitat for the plover, meaning that if there exists more visual obstruction by vegetation, then it would not be as suitable for the plovers to nest (Introduction, paragraph 3).

Distance to prairie dog colony edge is a continuous variable as it is a measurement value in metric units. As distance to prairie dog colony edge increases, the probability of Mountain Plover nest site use would decrease. This is also for similar reasons of soil disturbance and vegetation consumption by prairie dogs that make their colonies ideal habitats for plovers (Introduction, paragraph 3).

## c.

| Model number | Visual obstruction | Distance to prairie dog colony | Model description |
|-------------|:-----------:|:-----------:|:--------------------:|
| 0 |  |  | no predictors (null model) |
| 1 | X | X | all predictors (saturated model) |
| 2 | X |  | Visual obstruction |
| 3 |  | X | Distance to prairie dog colony |

## d.

```{r}
#| label: data cleaning mopl
mopl_clean <- mopl |> # create new clean dataframe for mopl
  clean_names() |> # clean column names to lowercase and underscores
  mutate(used_bin = recode_values(id, # create new column called used_bin to create 1s and 0s from id column
    "used" ~ "1", # when row says "used", use value "1"
    "available" ~ "0")) |> # when row says "available", use value "0"
  select(used_bin, edgedist, vo_nest) # select to display specific columns
```

```{r}
#| label: model fitting
# model 0: null model
model0 <- lm(used_bin ~ 1,
             data = mopl_clean)

# model 1: saturated model
model1 <- lm(used_bin ~ vo_nest + edgedist,
             data = mopl_clean)

# model 2: visual obstruction
model2 <- lm(used_bin ~ vo_nest,
             data = mopl_clean)
# model 3: distance to prairie dog colony
model3 <- lm(used_bin ~ edgedist,
             data = mopl_clean)
```

## e.

```{r}
#| label: AIC model selection
AICc(
  model0,
  model1,
  model2,
  model3) |> 
  arrange(AICc)
```

The best model as determined by Akaike's Information Criterion (AIC) is the one including the response of nest use (with the used_bin column) and visual obstruction as a predictor.

# Problem 4.

## a.
