---
title: "final_code"
author: "Sofia Perez"
date: "2026-03-16"
format: html
editor: visual
toc: true # use this to display a table of contents
---

github link:https://github.com/sofiaperez422/ENVS_193DS_winter-2026_final

## set up

```{r}
#| message: false
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(showtext) # to show custom text font on plots
font_add_google(name = "Montserrat", # name of font on Google font website
              family = "mont") # name to call in font as
showtext_auto()

kelp <- read_csv(here("data",
                      "Annual_Kelp_All_Years_20250903.csv")) # save annual-kelp data as object  kelp
mopl <- read_csv(here("data",
                      "MOPL_nest-site_selection_Duchardt_et_al._2019.csv")) # save MOPL-nest data as object MOPL
personal_data <- read_csv(here("data", 
                      "Personal_Data20260315.csv")) # save personal data as object
```

# Problem 1.

## a.

In part 1, they used Linear Regression. In part 2, they used Kruskal-Wallis rank sum test.

## b.

To accompany part 2, they could make a box plot, where the x-axis would be water year classification and the y-axis would be the flooded wetland area (m^2^ ). The visual components would show the individual distributions of each water year classification with quartiles that reveal the medians, as well as outliers that may exist as single points.

## c.

Part 1 – The change in flooded wetland area is not predicted by precipitation (linear regression, F(degrees of freedom) = F statistic, R^2^ = coefficient of determination, p = 0.11, $\alpha$ = significance level).

Part 2 – We found no difference ($\eta^2$ = effect size) in median flooded wetland area across water year classification (wet, above normal, below normal, dry, critical drought) (Kruskal-Wallis test, $\chi^2$(degrees of freedom) = H statistic, p = 0.12, $\alpha$ = significance level).

## d.

An additional variable that could influence the flooded area of a wetland is the daily Delta outflow, specifically applied as a median daily, which would be a continuous variable. This might influence measured wetland flooding area because it directly involves the activity of water managers intentionally moving water on and off flooded wetland units for reasons like leaching salts from soils and providing wetland habitat for breeding waterfowl around that time of year, which has potential to influence the overall area.

# Problem 2.

## a.

```{r}
kelp_clean <- kelp |> # create new clean dataframe titled kelp_clean from kelp
  
# function 3
clean_names() |>  # clean column names to lowercase and underscores

  # function 1
filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) |>  # filter out observations from 2000-09-22 with transect 6 and quad 40

# function 6
filter(site %in% c("IVEE", "NAPL", "CARP")) |>  # filter to only include sites "IVEE" "NAPL" and "CARP"

# function 8
mutate(site_name = case_when( # rename sites to full site names
  site == "NAPL" ~ "Naples",
  site == "IVEE" ~ "Isla Vista",
  site == "CARP" ~ "Carpinteria"
)) |> 

# function 2
mutate(site_name = as_factor(site_name), # convert site name entries to factors
       site_name = fct_relevel(site_name, # reorder sites using factor properties
                               "Naples", "Isla Vista", "Carpinteria")) |> 

# function 5
group_by(site_name, year) |>  # group by site and year

# function 7
summarize(mean_fronds = mean(fronds)) |>  # calculate mean of fronds

# function 4
ungroup() # remove grouping
```

```{r}
slice_sample(kelp_clean, # display 5 random rows from the sample
             n = 5) |> 
  str() # display the structure in a compact format
```

## b.

In the notes of the data set for that row it states "missing datasheet", and the number for the value of fronds in that "quad" in that "transect" is -99999. This value represents an "NA" value used when the data is either lost or not collected for a particular species (Methods and Protocols tab of the data page.)

## c.

```{r}
ggplot(data = kelp_clean,
       aes(x = year,
           y = mean_fronds,
           color = site_name)) +
  geom_line(linewidth = 0.8) +
  geom_point() +
  scale_color_manual(values = c("Isla Vista" = "bisque3", # change colors
                                "Naples" = "turquoise4", 
                                "Carpinteria" = "azure4")) +
  theme_minimal(base_family = "mont") +
  labs(x = "Year",
       y = "Mean kelp fronds per year",
       title = "Sites vary in mean kelp fronds per year",
       color = "Site") +
  theme(plot.title = element_text(hjust = 0.5))
```

## d. 

The site with the most variable mean kelp fronds per year is Isla Vista, as it has the highest points compared to its lowest points seen with the longest lines. The site with the least variable mean kelp fronds per year is Carpinteria, as it has the least peaking relative to the rest of its data points with shorter lines and is also the most consistent in the y-axis region that the points stay in.

# Problem 3.

## a.

# Problem 4.

## a.
